Dynamic Walking and Whole-Body Motion Planning for Humanoid Robots: an Integrated Approach

### 2012-12-05 paper.tex
Minor revision.

Section~\ref{sec:constraint-solver} describes a constraint solver,
Section~\ref{sec:goal-sampling} the goal manifold sampling, and
Section~\ref{sec:extension} the adaptation or RRT random extensions to
constrained motion planning.

\subsection{Multiple Constraints Solver}
\label{sec:constraint-solver}

@@

\textbf{In Algorithm~\ref{algo:newton} a Newton-Raphson method
  \cite{bonnans2006numerical}: starting from an initial value of $q$,
  $q$ is updated iteratively by $- \alpha \left(\frac{\partial
    f}{\partial q}(q)\right)^{+} f(q)$, where $\alpha$ denotes a gain
  and $\left(\frac{\partial f}{\partial q}(q)\right)^{+}$ denotes the
  Moore-Penrose pseudo-inverse of the Jacobian of $f(q)$. The use of
  an adaptive gain $\alpha$ that increases iteratively from an initial
  value $\alpha$ to a maximum value $\alpha_{max}$ has proven to avoid
  overshoots and accelerate convergence. The update rule relies on a
  real factor $w \in [0,1]$; the greater $w$ is, the faster $\alpha$
  will reach $\alpha_{max}$. Obviously, the solver convergence depends
  of the initial value of $q$, and a bad initialization can lead to
  either slow convergence or failure. A cutoff number of iterations
  $it_{max}$ is hence introduced to bypass these cases.}

\textbf{In this work, we observed that values of
  $\varepsilon=10^{-6}$, $\alpha=0.1$, $\alpha_{max}=0.95$ and $w=0.8$
  lead to good behavior, i.e. fast convergence and low failure rate.}

\begin{algorithm}
\caption{\texttt{constraint-solver}($q$, f, $\varepsilon$): find $q$ such that $f(q) = 0$}
\label{algo:newton}
\begin{algorithmic}
\STATE $i=0$
\WHILE{$\|f(q)\| > \varepsilon$ and $i\leq it_{max}$}
\STATE {\color{red} // $\left(.\right)^{+}$ denotes the Moore-Penrose pseudo-inverse}
\STATE $q \leftarrow$ $q - \alpha \left(\frac{\partial f}{\partial q}(q)\right)^{+} f(q)$
\STATE $i$ $\leftarrow$ $i+1$
\STATE {\color{red}// Make $\alpha$ tend toward $\alpha_{max}$}
\STATE $\alpha \rightarrow \alpha_{max} - w(\alpha_{max} - \alpha)$
\ENDWHILE
\IF {$\|f(q)\| \leq \varepsilon$}
\STATE return $q$
\ELSE
\STATE return failure
\ENDIF
\end{algorithmic}
\end{algorithm}

@@

The way we generate a goal configuration is the following:
\begin{enumerate}
\item \textbf{Shoot a random configuration $q_{rand}$ in $\mathcal{C}$ with
  uniform distribution.}
\item \textbf{Call \textit{constraint-solver}
  (Algorithm~\ref{algo:newton}) on $q_{rand}$, with $f(q)$ defined by
  the intersection of the planning and goal manifolds $\manifold \cap
  \goalmanifold$.}
\item If success, check for collisions.
\end{enumerate}

\textbf{Fig.~\ref{fig:goal} shows resulting random configurations
  enforcing balance ($\manifold$) and reaching ($\goalmanifold$)
  constraints for the HRP-2 robot.}

@@

The equivalent random  extension on a constrained manifold $\manifold$, 
defined by the constraint function $f$,
starts from a valid configuration $q_{near} \in \manifold$, and extends the tree
towards a random configuration $q_{rand}$, while keeping the constraints
satisfied. Extension attempts orthogonal to $\manifold$ are useless,
as newly added edges have to be included  in $\manifold$. To extend in directions
that follow the directions of $\manifold$, we rely on the framework of 
Jacobian-based prioritized inverse kinematics. Algorithm~\ref{alg:constrained} presents
the adaptation of the classic extend function, and Fig.~\ref{fig:gikrrt} illustrates this
extension. The idea is to first project $q_{rand}$ on the tangent
space to $\manifold$  at $q_{near}$. Let us call the projected configuration $q_{rand}'$.
$q_{rand}'$ is obtained by one step of prioritized inverse kinematics optimization. Let 
$q_{rand}''$  be the result of a call to $\texttt{constraint-solver}(q_{rand}',f,\epsilon)$.  It is 
the projection
of $q_{rand}'$ on $\manifold$. Instead of extending the tree from $q_{near}$ towards 
$q_{rand}$, the algorithm tries to extend from $q_{near}$ towards 
$q_{rand}''$\footnote{This presentation attempts to give a precise idea of the algorithm, without 
  focusing on technical details. Readers interested in the actual implementation can
  refer to \url{https://github.com/laas/hpp-constrained} and 
  \url{https://github.com/laas/hpp-constrained-planner}, where the corresponding open-source 
  code is available.}. While extending the tree,
the configurations along the new edge are automatically projected onto $\manifold$. These projections 
are not very costly if the edge is close to the constrained manifold.

@@

\begin{algorithm}[h]
  \caption{Constrained-Extend($\mathcal{T},q_{near},q_{rand},f,\epsilon$)}
  \label{alg:constrained}
  \begin{algorithmic}
    \STATE {\color{red} // $I$ denotes the identity matrix.}
    \STATE $q_{rand}' \leftarrow q_{near} +   
    \left(I -  
    \left(\frac{\partial f}{\partial q}(q_{near})\right)^{+}
    \frac{\partial f}{\partial q}(q_{near})\right) (q_{rand} - q_{near})$
    \IF {$\texttt{constraint-solver}(q_{rand}',f,\epsilon) = q_{rand}''$} 
    \STATE RRT::Extend($\mathcal{T},q_{near},q_{rand}''$)
    \ENDIF
  \end{algorithmic}
\end{algorithm}

@@

\textbf{We have run this set of motion planning problems 20 times;
  results are compiled in Table~\ref{table:reaching}. We have also
  measured the performance of \texttt{constraint-solver} (Algorithm
  \ref{algo:newton}) when used to project configurations on
  $\manifold$; the average number of iterations is 6.5 per call, and
  the success rate, i.e. the ratio of the number of successfully
  projected configurations over the total number of calls, is above 95
  percent. This success rate, high as it is, could be further improved
  by sampling a better initial configuration of $\mathcal{C}$, for
  example by introducing a heuristic bias towards statically balanced
  configurations. Nevertheless we choose to sample $\mathcal{C}$
  uniformly-randomly for the sake of genericity.}

>>>>

Section~\ref{sec:SolveConstraints} describes a constraint solver,
Section~\ref{sec:goal-sampling} the goal manifold sampling, and
Section~\ref{sec:extension} the adaptation or RRT random extensions to
constrained motion planning.

\subsection{Constraint Solver}
\label{sec:SolveConstraints}

@@

\textbf{Algorithm~\ref{algo:newton} implements a Newton-Raphson method
  \cite{bonnans2006numerical}: starting from an initial value of $q$,
  $q$ is updated iteratively by $- \alpha \left(\frac{\partial
    f}{\partial q}(q)\right)^{+} f(q)$, where $\alpha$ denotes a gain
  and $\left(\frac{\partial f}{\partial q}(q)\right)^{+}$ denotes the
  Moore-Penrose pseudo-inverse of the Jacobian of $f(q)$. The use of
  an adaptive gain $\alpha$, which increases iteratively from an
  initial value $\alpha$ to a maximum value $\alpha_{max}$, allows the
  overshoot avoidance and convergence acceleration. The update rule
  relies on a real factor $w \in [0,1]$; the greater $w$ is, the
  faster $\alpha$ will reach $\alpha_{max}$. Obviously, the solver
  convergence depends on the initial value of $q$, and a bad
  initialization can lead to either slow convergence or failure. A
  cutoff number of iterations $it_{max}$ is hence introduced to bypass
  these cases.}

\textbf{We observed that values of $\varepsilon=10^{-6}$,
  $\alpha=0.1$, $\alpha_{max}=0.95$ and $w=0.8$ lead to good behavior,
  i.e. fast convergence and low failure rate. These values are kept
  constant for all scenarios in this work.}

\begin{algorithm}
\caption{\texttt{SolveConstraints}($q$, f, $\varepsilon$): find $q$ such that $f(q) = 0$}
\label{algo:newton}
\begin{algorithmic}
\STATE $i=0$
\WHILE{$\|f(q)\| > \varepsilon$ and $i\leq it_{max}$}
\STATE {\color{red} // $\left(.\right)^{+}$ denotes the Moore-Penrose pseudo-inverse}
\STATE $q \leftarrow$ $q - \alpha \left(\frac{\partial f}{\partial q}(q)\right)^{+} f(q)$
\STATE $i$ $\leftarrow$ $i+1$
\STATE {\color{red}// Make $\alpha$ tend toward $\alpha_{max}$}
\STATE $\alpha \rightarrow \alpha_{max} - w(\alpha_{max} - \alpha)$
\ENDWHILE
\IF {$\|f(q)\| \leq \varepsilon$}
\STATE return $q$
\ELSE
\STATE return failure
\ENDIF
\end{algorithmic}
\end{algorithm}

@@

The way we generate a goal configuration is the following:
\begin{enumerate}
\item \textbf{Shoot a random configuration $q_{rand}$ in $\mathcal{C}$ with
  uniform distribution.}
\item \textbf{Call SolveConstraints (Algorithm~\ref{algo:newton}) on
  $q_{rand}$, with $f(q)$ defined by the intersection of the planning
  and goal manifolds $\manifold \cap \goalmanifold$.}
\item If success, check for collisions.
\end{enumerate}

\textbf{Fig.~\ref{fig:goal} shows resulting random configurations
  which satisfy both balance ($\manifold$) and reaching
  ($\goalmanifold$) constraints for the HRP-2 robot.}

@@

The equivalent random  extension on a constrained manifold $\manifold$, 
defined by the constraint function $f$,
starts from a valid configuration $q_{near} \in \manifold$, and extends the tree
towards a random configuration $q_{rand}$, while keeping the constraints
satisfied. Extension attempts orthogonal to $\manifold$ are useless,
as newly added edges have to be included  in $\manifold$. To extend in directions
that follow the directions of $\manifold$, we rely on the framework of 
Jacobian-based prioritized inverse kinematics. Algorithm~\ref{alg:constrained} presents
the adaptation of the classic extend function, and Fig.~\ref{fig:gikrrt} illustrates this
extension. The idea is to first project $q_{rand}$ on the tangent
space to $\manifold$  at $q_{near}$. Let us call the projected configuration $q_{rand}'$.
$q_{rand}'$ is obtained by one step of prioritized inverse kinematics optimization. Let 
$q_{rand}''$  be the result of a call to $\texttt{SolveConstraints}(q_{rand}',f,\epsilon)$.  It is 
the projection
of $q_{rand}'$ on $\manifold$. Instead of extending the tree from $q_{near}$ towards 
$q_{rand}$, the algorithm tries to extend from $q_{near}$ towards 
$q_{rand}''$\footnote{This presentation attempts to give a precise idea of the algorithm, without 
  focusing on technical details. Readers interested in the actual implementation can
  refer to \url{https://github.com/laas/hpp-constrained} and 
  \url{https://github.com/laas/hpp-constrained-planner}, where the corresponding open-source 
  code is available.}. While extending the tree,
the configurations along the new edge are automatically projected onto $\manifold$. These projections 
are not very costly if the edge is close to the constrained manifold.

@@

\begin{algorithm}[h]
  \caption{ConstrainedExtend($\mathcal{T},q_{near},q_{rand},f,\epsilon$)}
  \label{alg:constrained}
  \begin{algorithmic}
    \STATE {\color{red} // $I$ denotes the identity matrix.}
    \STATE $q_{rand}' \leftarrow q_{near} +   
    \left(I -  
    \left(\frac{\partial f}{\partial q}(q_{near})\right)^{+}
    \frac{\partial f}{\partial q}(q_{near})\right) (q_{rand} - q_{near})$
    \IF {$\texttt{SolveConstraints}(q_{rand}',f,\epsilon) = q_{rand}''$}
    \STATE RRT::Extend($\mathcal{T},q_{near},q_{rand}''$)
    \ENDIF
  \end{algorithmic}
\end{algorithm}

@@

\textbf{We have run this set of motion planning problems 20 times;
  results are compiled in Table~\ref{table:reaching}. We have also
  measured the performance of \texttt{SolveConstraints} (Algorithm
  \ref{algo:newton}) when used to project configurations on
  $\manifold$; the average number of iterations is 6.5 per call, and
  the success rate, i.e. the ratio of the number of successfully
  projected configurations over the total number of calls, is above 95
  percent. This success rate, high as it is, could be further improved
  by sampling a better initial configuration of $\mathcal{C}$, for
  example by introducing a heuristic bias towards statically balanced
  configurations. Nevertheless we choose to sample $\mathcal{C}$
  uniformly-randomly for the sake of genericity.}
